Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-03-02 Paper/Manuscript_Stats&Comp/reply_letter_STCO.tex
update response letter

%Paul: all responses will be send to all reviewers anyway
%Authors would like the summary of the updates and insights presented above to be sent to all of the reviewers.

@@

@@

%Paul: This list is way too long and detailed and appears to be redundant at times. I recommend focusing on between 4 and 7 or so points but more will be too much.

@@

{\color{red} (TODO: I've not done this yet, but I could do it before the deadline.)}

@@

\item A potential fitting overestimating effect when using too many basis functions relative to the wigglyness of the function to be learned has been commented. {\color{red}

(TODO: Currently we just have outlined but not discussed this. If we had time, it would be nice to discover why this overestimating effect is happening.)}

@@

%Paul: Don't forget the thank the reviewers for their comments to show some more appreciation of their comments and time they put into the review.

@@

%Paul: The statement below is incorrect. I have not implemented the diagnostic procedure in brms.
%The diagnostic tool which consists in iterative steps to fit and diagnose the model has also been implemented in the \textit{brms} R-package. 
%This diagnostic iterative procedure is fundamentally based on a diagnostic rule to check whether the approximation is accurate and numerical equations allowing updating the key parameters of the method.

@@

%Paul: I recommend highlighting either the reviewers comments or our responses in bold to clearly differentiate what the reviewers wrote and what we wrote. Example below.

@@

Reviewer \#1 comments "The results are intuitive, but the process of determining the parameters in practice remains somewhat cumbersome as one needs to inspect the curves in Figures 6 and 7 to make the decision and there are no definite guidelines on how to modify the parameters when observing insufficient accuracy (e.g. no hints on how much one should increase m)."

@@

Reviewer \#1 comments "Furthermore, to use the approximation with any other kernel besides square exponential or Matérn one would essentially need to repeat the same analysis from scratch. This limits the value in probabilistic programming context. The authors indicate existing GP  libraries are limited and favour Stan for its generality, but since we now only have support for two specific kernels one could easily argue this approach is limited as well."

@@

Reviewer \#1 comments "it still feels like many educated practitioners would have found similar values for m and c by following a conventional validation procedure of applying the approximation with range of values for both on some example functions. While they would probably have ended up using slightly too large m or might spend a bit more computation while finding the best values, the practical difference might not be that big."

@@

Reviewer \#1 comments "The case studies are interesting and demonstrate the approximation scheme works in practice for simple statistical models while providing clear example codes, but they could be more focused towards the main goal of the paper."

@@

{\color{red} (TODO: I've not done this yet, but I could do it before the deadline.)}
\end{itemize}

@@

Reviewer \#1 comments "Section 5.1 takes approximately four full pages to present results for a simple artificial data. It describes one concrete example of setting the parameters (that remains somewhat vague -- "In the first iteration, we choose...", "In the second iteration, m is increased to..."), but the detailed comparisons against splines could have been much more brief in a paper focusing on how to make the approximation easier to use in practice."

@@

Reviewer \#1 comments "Section 5.2 is similarly a bit besides the point -- the example model is interesting but regarding the choice of the parameters you simply say "We use m=30 basis functions and a boundary factor c=1.5" followed by stating that the length-scale estimate is good. This is what we would expect to see in a paper that later uses the method, but for a paper that is introducing the diagnostic I would expect more details and e.g. explanations of what would happen if the parameters were poorly chosen. I understand fully that verbally describing an iterative process is difficult, but was still a bit let down by these case studies."

@@

Reviewer \#1 comments "Section 5.3 presents a multivariate case and seems to mostly provide a negative result on applicability of HSGP for multivariate cases. This is fine as the method is valuable even if limited to low dimensions, but it would have been more interesting to see a more detailed analysis of the intermediate region, especially D=2 and D=3 that are frequent in spatial and spatiotemporal analysis problems -- a practitioner would appreciate clear guidelines on when to use HSGP and how to recognise when it is not a good solution."

@@

We have added a new case study to Section 5 which consists in a wiggly 2D simulated data set. We have applied the diagnostic procedure on it. Even though the diagnostic was built for unidimensional kernels, we demonstrate its applicability and usefulness in multidimensional cases by diagnosing individually the different dimensions in each iteration. It allows finding the  optimal solution with the minimum requirements that can be essential to use the HSGP model in multidimensional cases. We have included in this new case study a brief discussion and presentation of additional results concerning computational cost in 2D, 3D and 4D, from case studies initially included in the online material. {\color{red} (TODO: I've not done this yet, but I could do it before the deadline.)}

@@

Reviewer \#1 comments "Overall, I believe the paper has clear value for practical use and is worth publishing in some form: It provides nice code examples for HSGP via case studies and an iterative procedure for diagnosing when the approximation accuracy is good enough. However, the scientific novelty of the work is fairly limited and perhaps insufficient for publication in a top venue, in particular because there are no claims or contributions that would extend beyond this specific approximation scheme. There are also no clear scientific insights in the process itself, since the analysis is purely empirical and follows a straightforward protocol. In summary, the manuscript is a borderline case even if addressing the remarks below."

@@

\#1.1 - Reviewer \#1 comments: "The diagnostic involves comparing the estimated length-scale $\hat l$ against a minimum $l^*$ determined by m and c. I have trouble understanding the full validity of this process. I agree with your statement on page 8 about $\hat l$ being accurate if the approximation is accurate, but you do not state anything definite about the quality of $\hat l$ when the approximation is inaccurate. However, the process seems to implicitly assume we can still rely on $\hat l$ even then, with very relaxed justification ("If $\hat l$ does not exceed ... the approximation *may* be inaccurate"). Am I correct that your method would fail to diagnose a bad approximation if $\hat l$ was overestimated? I would like to see this discussion extended, at least by explicitly mentioning possible failure modes."

@@

\noindent \#1.2 - Reviewer \#1 comments: "To me it seems Fig 5 (left) already hints that you tend to underestimate l when using bad approximation and the procedure may be valid in practice, but this is not transparent in the paper."

@@

\noindent \#1.3 Reviewer \#1 comments: "Do we have a reason for the error to be always towards understimating the lengthscale? Could this be proven?"

@@

Reviewer \#1 comments: "The process involves checking Figure 6 (or 7) for verifying whether the approximation is valid. This feels cumbersome in practice, and for better PP integration I would expect proper software support for this. It would be nice to at least outline how this could work, but even better to provide an actual tool. Both a visual tool plotting the current estimate as in Fig 13 and a textual feedback like "$\hat l$ is too small, you should either increase m or decrease c a bit -- we recommend the former" could work. One could also make this completely automatic, a loop that keeps on modifying the parameters until the threshold is reached. I would like you to discuss which of these options you would prefer and why."

@@

Reviewer \#1 comments: "The paper is in general well written, but the structure could be improved. I think the specific diagnostic is now lost in the middle of the story in Sections 4 and 5 and would benefit from clear presentation in dedicated section (that would also be a good place to discuss the software support; see comment 2 above). The writing of both Sections 4 and 5 is also quite verbose and text-heavy, which makes reading the paper somewhat tedious. Section 4.1 is 200+ lines of text with no structure and as mentioned above both 5.1 and 5.2 are somewhat long compared to their value. Since you anyway provide exact code and re-use models from previous papers, you could compress the general discussion of the models and focus more on the process of determining the approximation parameters."

@@

Reviewer \#1 comments: " It would be really interesting to see a bit more about how HSGP works for functions with 2-3 inputs, and in particular about how a practitioner can recognise whether it makes sense to use HSGP for their case or not. To me it seems clear HSGP is very often a reasonable method for D=1 and almost certainly breaks down for D>=5 or so unless working with very smooth functions (and for them almost any approximation works well), but clear diagnostics for quickly determining whether it could work for D=2 or D=3 would increase the contribution notably. Now the practical process would be to try it out and keep on increasing m until it starts working or one notices the computation is too slow, which feels wasteful -- as a practitioner I would rather just skip HSGP completely if finding out it does not help takes longer than solving the problem without the approximation."

@@

Reviewer \#2 comments "The paper gives experimental results for the method of Solin and Sarkka  (2020). Overall the idea such experiments is interesting, but the actual experiments are not extensive enough given the pure applied nature of this work."

@@

Reviewer \#2 comments "Experiments are mainly low dimensional (all 1D except a single 4D example). Please discuss high dimensional cases ; indeed equation 8 suggests poor scaling in dimension. Can it be applied to high dimensional data?".

@@

We have added a new case study to Section 5 which consists in a wiggly 2D simulated data set. We have applied the diagnostic procedure on it. Even though the diagnostic was built for unidimensional kernels, we demonstrate its applicability and usefulness in multidimensional cases by diagnosing individually the different dimensions in each iteration. It allows finding the  optimal solution with the minimum requirements that can be essential to use the HSGP model in multidimensional cases. We have included in this new case study a brief discussion and presentation of additional results concerning computational cost in 2D, 3D and 4D, from case studies initially included in the online material. {\color{red} (TODO: I've not done this yet, but I could do it before the deadline.)}

@@

\subsubsection*{R\#2 Comment \#3}
Reviewer \#2 comments "There is no comparison with other techniques (except a spline method) ! Please compare with, e.g. "MCMC for Variationally Sparse Gaussian Processes" ; although this is SVI, it appears directly applicable for the problems considered here. Please also compare with non MCMC variational methods - while not directly applicable to generic MCMC methods, these are the sttandard and it would be nice to see how they compare in runtime."

@@

Reviewer \#2 comments "On the eigenfunction expansion: 1) note that this is very similar to the Mercer expansion of e.g. Walder et al, "Fast Bayesian Intensity Estimation for the Permanental Process". 2) the Nystrom method also yields an approximate eigen expansion which has pros and cons to the method of the submission. For example, the Nystrom applies to any kernel (not just stationary) and yields a linear GP approximation, but requires some overhead for the setup. Given this submission is purely experimental, it would be nice to see comparisons.".

>>>>

@@

{\color{blue} Paul: This list is way too long and detailed and appears to be redundant at times. I recommend focusing on between 4 and 7 or so points but more will be too much.}

@@

@@

{\color{blue} (Gabriel: I've not done this yet, but I could do it before the deadline.)}

@@

\item A potential fitting overestimating effect when using too many basis functions relative to the wigglyness of the function to be learned has been commented.

{\color{blue} (Gabriel: Currently we just have outlined but not discussed this. If we had time, it would be nice to discover why this overestimating effect is happening.)}

@@

Authors are really grateful to the reviewer for their valuable comments which have been very helpful to improve the paper.

@@

We are planning to implement the diagnostic tool which consists in iterative steps to fit and diagnose the model in the \textit{brms} R-package.

%This diagnostic iterative procedure is fundamentally based on a diagnostic rule to check whether the approximation is accurate and numerical equations allowing updating the key parameters of the method.

@@

@@

\textbf{The results are intuitive, but the process of determining the parameters in practice remains somewhat cumbersome as one needs to inspect the curves in Figures 6 and 7 to make the decision and there are no definite guidelines on how to modify the parameters when observing insufficient accuracy (e.g. no hints on how much one should increase m).}

@@

\textbf{Furthermore, to use the approximation with any other kernel besides square exponential or Matérn one would essentially need to repeat the same analysis from scratch. This limits the value in probabilistic programming context. The authors indicate existing GP  libraries are limited and favour Stan for its generality, but since we now only have support for two specific kernels one could easily argue this approach is limited as well.}

@@

\textbf{it still feels like many educated practitioners would have found similar values for m and c by following a conventional validation procedure of applying the approximation with range of values for both on some example functions. While they would probably have ended up using slightly too large m or might spend a bit more computation while finding the best values, the practical difference might not be that big.}

@@

\textbf{The case studies are interesting and demonstrate the approximation scheme works in practice for simple statistical models while providing clear example codes, but they could be more focused towards the main goal of the paper.}

@@

{\color{red} (Gabriel: I've not done this yet, but I could do it before the deadline.)}
\end{itemize}

@@

\textbf{Section 5.1 takes approximately four full pages to present results for a simple artificial data. It describes one concrete example of setting the parameters (that remains somewhat vague -- "In the first iteration, we choose...", "In the second iteration, m is increased to..."), but the detailed comparisons against splines could have been much more brief in a paper focusing on how to make the approximation easier to use in practice.}

@@

\textbf{Section 5.2 is similarly a bit besides the point -- the example model is interesting but regarding the choice of the parameters you simply say "We use m=30 basis functions and a boundary factor c=1.5" followed by stating that the length-scale estimate is good. This is what we would expect to see in a paper that later uses the method, but for a paper that is introducing the diagnostic I would expect more details and e.g. explanations of what would happen if the parameters were poorly chosen. I understand fully that verbally describing an iterative process is difficult, but was still a bit let down by these case studies.}

@@

\textbf{Section 5.3 presents a multivariate case and seems to mostly provide a negative result on applicability of HSGP for multivariate cases. This is fine as the method is valuable even if limited to low dimensions, but it would have been more interesting to see a more detailed analysis of the intermediate region, especially D=2 and D=3 that are frequent in spatial and spatiotemporal analysis problems -- a practitioner would appreciate clear guidelines on when to use HSGP and how to recognise when it is not a good solution.}

@@

We have added a new case study to Section 5 which consists in a wiggly 2D simulated data set. We have applied the diagnostic procedure on it. Even though the diagnostic was built for unidimensional kernels, we demonstrate its applicability and usefulness in multidimensional cases by diagnosing individually the different dimensions in each iteration. It allows finding the  optimal solution with the minimum requirements that can be essential to use the HSGP model in multidimensional cases. We have included in this new case study a brief discussion and presentation of additional results concerning computational cost in 2D, 3D and 4D, from case studies initially included in the online material. {\color{red} (Gabriel: I've not done this yet, but I could do it before the deadline.)}

@@

\textbf{Overall, I believe the paper has clear value for practical use and is worth publishing in some form: It provides nice code examples for HSGP via case studies and an iterative procedure for diagnosing when the approximation accuracy is good enough. However, the scientific novelty of the work is fairly limited and perhaps insufficient for publication in a top venue, in particular because there are no claims or contributions that would extend beyond this specific approximation scheme. There are also no clear scientific insights in the process itself, since the analysis is purely empirical and follows a straightforward protocol. In summary, the manuscript is a borderline case even if addressing the remarks below.}

@@

\textbf{\#1.1 - The diagnostic involves comparing the estimated length-scale $\hat l$ against a minimum $l^*$ determined by m and c. I have trouble understanding the full validity of this process. I agree with your statement on page 8 about $\hat l$ being accurate if the approximation is accurate, but you do not state anything definite about the quality of $\hat l$ when the approximation is inaccurate. However, the process seems to implicitly assume we can still rely on $\hat l$ even then, with very relaxed justification ("If $\hat l$ does not exceed ... the approximation *may* be inaccurate"). Am I correct that your method would fail to diagnose a bad approximation if $\hat l$ was overestimated? I would like to see this discussion extended, at least by explicitly mentioning possible failure modes.}

@@

\textbf{\#1.2 - To me it seems Fig 5 (left) already hints that you tend to underestimate l when using bad approximation and the procedure may be valid in practice, but this is not transparent in the paper.}

@@

\textbf{\#1.3 - Do we have a reason for the error to be always towards understimating the lengthscale? Could this be proven?}

@@

\textbf{The process involves checking Figure 6 (or 7) for verifying whether the approximation is valid. This feels cumbersome in practice, and for better PP integration I would expect proper software support for this. It would be nice to at least outline how this could work, but even better to provide an actual tool. Both a visual tool plotting the current estimate as in Fig 13 and a textual feedback like "$\hat l$ is too small, you should either increase m or decrease c a bit -- we recommend the former" could work. One could also make this completely automatic, a loop that keeps on modifying the parameters until the threshold is reached. I would like you to discuss which of these options you would prefer and why.}

@@

\textbf{The paper is in general well written, but the structure could be improved. I think the specific diagnostic is now lost in the middle of the story in Sections 4 and 5 and would benefit from clear presentation in dedicated section (that would also be a good place to discuss the software support; see comment 2 above). The writing of both Sections 4 and 5 is also quite verbose and text-heavy, which makes reading the paper somewhat tedious. Section 4.1 is 200+ lines of text with no structure and as mentioned above both 5.1 and 5.2 are somewhat long compared to their value. Since you anyway provide exact code and re-use models from previous papers, you could compress the general discussion of the models and focus more on the process of determining the approximation parameters.}

@@

\textbf{It would be really interesting to see a bit more about how HSGP works for functions with 2-3 inputs, and in particular about how a practitioner can recognise whether it makes sense to use HSGP for their case or not. To me it seems clear HSGP is very often a reasonable method for D=1 and almost certainly breaks down for D>=5 or so unless working with very smooth functions (and for them almost any approximation works well), but clear diagnostics for quickly determining whether it could work for D=2 or D=3 would increase the contribution notably. Now the practical process would be to try it out and keep on increasing m until it starts working or one notices the computation is too slow, which feels wasteful -- as a practitioner I would rather just skip HSGP completely if finding out it does not help takes longer than solving the problem without the approximation.}

@@

\textbf{The paper gives experimental results for the method of Solin and Sarkka  (2020). Overall the idea such experiments is interesting, but the actual experiments are not extensive enough given the pure applied nature of this work.}

@@

\textbf{Experiments are mainly low dimensional (all 1D except a single 4D example). Please discuss high dimensional cases ; indeed equation 8 suggests poor scaling in dimension. Can it be applied to high dimensional data?.}

@@

We have added a new case study to Section 5 which consists in a wiggly 2D simulated data set. We have applied the diagnostic procedure on it. Even though the diagnostic was built for unidimensional kernels, we demonstrate its applicability and usefulness in multidimensional cases by diagnosing individually the different dimensions in each iteration. It allows finding the  optimal solution with the minimum requirements that can be essential to use the HSGP model in multidimensional cases. We have included in this new case study a brief discussion and presentation of additional results concerning computational cost in 2D, 3D and 4D, from case studies initially included in the online material. {\color{blue} (Gabriel: I've not done this yet, but I could do it before the deadline.)}

@@

\subsubsection*{R\#2 Comment \#3}

\textbf{There is no comparison with other techniques (except a spline method) ! Please compare with, e.g. "MCMC for Variationally Sparse Gaussian Processes" ; although this is SVI, it appears directly applicable for the problems considered here. Please also compare with non MCMC variational methods - while not directly applicable to generic MCMC methods, these are the sttandard and it would be nice to see how they compare in runtime.}

@@

\textbf{On the eigenfunction expansion: 1) note that this is very similar to the Mercer expansion of e.g. Walder et al, "Fast Bayesian Intensity Estimation for the Permanental Process". 2) the Nystrom method also yields an approximate eigen expansion which has pros and cons to the method of the submission. For example, the Nystrom applies to any kernel (not just stationary) and yields a linear GP approximation, but requires some overhead for the setup. Given this submission is purely experimental, it would be nice to see comparisons.}
